---
title: "California Oil Spill Incident Tracking"
description: |
  The California Office of Spill Prevention and Response (OSPR) Incident Tracking Database tracks and records oil spills throughout the state. The analysis in this report utilized the database to visualize findings with California county shapefile data. For both shapefile datasets the CRS is WGS84 (epsg: 3857) so no transformation was needed.
author:
  - name: Zoe Duerksen-Salm
date: 02-23-2021
output:
  distill::distill_article:
    self_contained: false
    code_folding: hide
---

```{r setup, include=FALSE, messae=FALSE, warnings=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(here)
library(broom)
library(janitor)
library(sf)
library(tmap)
library(tidyverse)
library(dplyr)
```

```{r}
# Read in data from CA DFW Oil Spill Incident Tracking
osi_track <- read_sf(here("data","OSI_tracking","OSI_tracking.shp")) %>% 
  clean_names()

# Read in CA County shapefile data from the US Census Bureau
ca_counties <- read_sf(here("data","ca_counties","CA_Counties_TIGER2016.shp")) %>% 
  clean_names()

# Subset CA county shapefile data to include necessary variables
ca_subset <- ca_counties %>% 
  select(name, aland) %>% 
  rename(county_name = name, land_area = aland)

# Rename `inlandmaari` variable for visualization purposes
osi_subset <- osi_track %>% 
  rename("Incident Location" = inlandmari)
```

## Chloropleth Map

### California Inland Oil Spill Events in 2008
```{r}
# Combine osi and county shapefile datasets
ca_osi <- ca_subset %>% 
  st_join(osi_track)

# Filter to include only `inland` counts
osi_counts <- ca_osi %>% 
  filter(inlandmari == "Inland") %>% 
  count(localecoun) %>% 
  rename("Incidents" = n)

# Create visualization
ggplot() +
  geom_sf(data = ca_subset, fill = "grey97", color = "grey50", size = 0.2)+
  geom_sf(data = osi_counts, aes(fill = Incidents), color = "grey50", size = 0.2) +
  scale_fill_gradientn(colors = c("lightgoldenrodyellow","orange", "red3")) +
  theme_minimal()
```
**Total counts for oil spill events in inland California for the year 2008** visualized as a static chloropleth map. Total counts were calculated by county and are shown with color. Higher counts are visualized as darker colors, lighter colors show lower counts.

---

## Data Source

**Oil Spill Incident Tracking [ds394] (2008).** [Digital map](https://map.dfg.ca.gov/metadata/ds0394.html). California Department of Fish and Game, Office of Spill Prevention and Response.